The Design and Implementation of the Synchronous Language \CEU

### 2016-04-11 paper.tex
...

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with simpler semantics and 
more fine-grained control over program execution.
%
\CEU adopts an event-triggered notion of time that enables compile-time 
temporal analysis to detect conflicting statements, resulting in deterministic 
and concurrency-safe programs.
%
Using Esterel as a base, we discuss the particularities of our design, such as 
stack-based internal events, the temporal analysis, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
% TODO: simple simple
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution, and is amenable to a temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C, which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and device drivers.
%
We evaluated the expressiveness of \CEU in comparison to hand-crafted 
event-driven code in \C and attested a reduction in source code size (around 
25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes which can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

\CEU is a synchronous reactive language inspired by Esterel in which programs 
advance in a sequence of discrete reactions to external events.
%
%Regarding the similarities with Esterel,
%\CEU has a strong imperative flavor, with explicit control flow through 
%sequences, loops, and parallel compositions.
%
Like Esterel, \CEU is designed for control-intensive applications, supporting 
concurrent lines of execution, known as \emph{trails}, and broadcast 
communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this discrete notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

Figure~\ref{lst.abro} illustrates the syntactic similarities between the 
languages, showing side-by-side the implementations in Esterel \ax and \CEU \bx 
for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 4--11) which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

.
\end{lstlisting}
\centering\small{\ax Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
input  void A, B;
output void O;
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\centering\small{\bx \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

@@

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish with a summary of our design
    (\ref{sec.ceu.summary}).
%
We present the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events from the environment.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute, one after the other, until they await or 
      terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes a negligible time on step 
2 and is always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running, it is enqueued to occur in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} shows event occurrences over a continuous timeline.
The other boxes show discrete logical units of time in which the same occurring 
events fit in different ways.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks},
in which reactions must fit~\cite{esterel.fixed}.
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} is considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}, which are adjusted to the CPU time 
the corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two reactions 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), 
Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
though \code{A} and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to detect simultaneous button clicks in 
\CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events in \CEU never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about concurrency and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction of a given event.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulates for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO-no: Such analsysis is orthogonal

A limitation of event-triggered execution is that all program behavior must be 
purely reactive, given that no code can execute in the absence of inputs.
%
Tick-triggered execution allows for active behavior, since code can execute 
regulary on every tick.
%
Although \CEU supports asynchronous active execution~\cite{ceu.tr}, we still 
consider it as purely reactive since its synchronous core cannot express active 
behavior.

@@

Esterel makes no semantic distinctions between internal and external signals.
%In particular, programs can emit external and internal signals simultaneously, 
%with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
%In \CEU, subsequent external events coming from the environment define a 
%timeline, which programs cannot interfere by emitting new external events.
%
%To cope with deterministic intra-reaction communication, \CEU supports 
%internal events which programs can \code{emit} and \code{await}.
In \CEU, in contrast with queue-based external events, internal events follow a 
stack-based execution policy similar to subroutine calls in typical programming 
languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order (ln. 6,9).
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates (ln. 5--6).
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events provide a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence, the calling trail resumes and passes the assertion 
test (ln. 11).

\CEU also supports nested \code{emit} invocations for internal events.
For instance, the code in Figure~\ref{lst.sub} that executes after the 
\code{await inc} (ln.  4) could \code{emit} another event, creating a new level 
in the stack.
The runtime stack constitutes fine-grained micro reactions, one on top of the 
other, all inside the same reaction to an external event.
%Internal events provide a fine-grained intra-reaction control mechanism.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
with the caller (ln. 8--11).
\label{lst.sub}
}
\end{figure}

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself is always ignored, 
given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring data stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, similarly to coroutines~\cite{lua.coroutines}.
%
In a previous work, we build on top of internal events other advanced control 
mechanisms, such as resumable exceptions and reactive 
variables~\cite{ceu.rem13}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.

@@

Esterel is only deterministic with respect to external behavior: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages have intra-reaction determinism 
by enforcing an arbitrary execution order for statements in multiple lines of 
execution (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

>>>>

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
more fine-grained control over program execution.
%
\CEU uses an event-triggered notion of time that enables compile-time temporal 
analysis to detect conflicting statements, resulting in deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we present the particularities of our design, such as 
stack-based internal events, the temporal analysis, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU has a simple semantics with fine-grained control for intra-reaction 
execution, and is amenable to a temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C, which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and device drivers.
%
We evaluated the expressiveness of \CEU in comparison to hand-crafted 
event-driven code in \C and attested a reduction in source code size (around 
25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes and can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

\CEU is a synchronous reactive language inspired by Esterel in which programs 
advance in a sequence of discrete reactions to external events.
%
%Regarding the similarities with Esterel,
%\CEU has a strong imperative flavor, with explicit control flow through 
%sequences, loops, and parallel compositions.
%
Like Esterel, \CEU is designed for control-intensive applications, supporting 
concurrent lines of execution, known as \emph{trails}, and broadcast 
communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this discrete notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with (core) Esterel and synchronous languages 
in general~\cite{esterel.preemption}:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

Figure~\ref{lst.abro} illustrates the syntactic similarities between the 
languages, showing side-by-side the implementations in Esterel \ax and \CEU \bx 
for the following control specification:
%
\emph{``Emit an output O as soon as inputs A and B occur.
Reset this behavior each time input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 4--11) which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

.
\end{lstlisting}
\centering\small{\ax Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
input  void A, B;
output void O;
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\centering\small{\bx \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
A \code{par/and} terminates when both trails in parallel terminate.
A \code{par/or}  terminates when any trail terminates, aborting the 
other.
\label{lst.abro}
}
\end{figure}

@@

In the subsections that follow, we discuss the main differences from \CEU to 
Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish the section with a summary of our design
    (\ref{sec.ceu.summary}).
%
We present the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events from the environment.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate, one after the other.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes a negligible time on step 
2 and is always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running, it is enqueued to occur in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} shows event occurrences over a continuous timeline.
The other boxes show how the same occurring events fit differently in each 
logical notion of time.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1} and invades 
\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods because it adjusts the tick lengths to match the CPU times 
for the reactions.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} at \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
behavior in \CEU is similar to \code{Box-2} for its first two reactions 
(\code{tick-0} and \code{tick-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
%
Note that, by definition, each reaction corresponds to an event occurrence of 
the same length.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we consider that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
though \code{A} and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to detect simultaneous button clicks in 
\CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events in \CEU never overlap because they are atomic.
Automatic mutual exclusion simplifies the reasoning about concurrency and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction of a given event.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulates for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, requiring \emph{worst case reaction time} 
analysis to infer appropriate tick lengths~\cite{esterel.fixed}.
% TODO-no: Such analsysis is orthogonal

A limitation of event-triggered execution is that all program behavior is 
purely reactive, given that no code can execute in the absence of inputs.
%
Tick-triggered execution allows for active behavior, since code can execute 
regulary on every tick.
%
Although \CEU supports active \emph{asynchronous} execution~\cite{ceu.tr}, its 
synchronous core is still purely reactive.

@@

In Esterel, the behavior of internal and external signals is equivalent.
%In particular, programs can emit external and internal signals simultaneously, 
%with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
%In \CEU, subsequent external events coming from the environment define a 
%timeline, which programs cannot interfere by emitting new external events.
%
%To cope with deterministic intra-reaction communication, \CEU supports 
%internal events which programs can \code{emit} and \code{await}.
In \CEU, in contrast with queue-based external events, internal events follow a 
stack-based execution policy similar to subroutine calls in typical programming 
languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, the program emits \code{B} (ln.  4--5) and 
both events become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order (ln. 6,9).
%
In \CEU, when \code{A} occurs, the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates (ln. 5--6).
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events provide fine-grained execution control and can express a 
limited form of subroutines, as depicted in Figure~\ref{lst.sub}.
The ``subroutine'' \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine through an \code{emit 
inc} (ln. 10) in reaction to some code (ln. 9).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence, the calling trail resumes and passes the assertion 
test (ln. 10--11).

\CEU also supports nested \code{emit} invocations for internal events.
For instance, the body of the subroutine \code{inc} in Figure~\ref{lst.sub} 
could \code{emit} another event after awaking (ln. 4), creating a new level in 
the stack.
The runtime stack constitutes fine-grained micro reactions, with one on top of 
the other, all inside the same reaction to an external event.
%Internal events provide a fine-grained intra-reaction control mechanism.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    <...>
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
with the caller (ln. 8--11).
\label{lst.sub}
}
\end{figure}

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself is always ignored, 
given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring data stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, similarly to coroutines~\cite{lua.coroutines}.
%
In a previous work, we build other advanced control mechanisms on top of 
internal events, such as resumable exceptions and reactive 
variables~\cite{ceu.rem13}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.

@@

Esterel is only deterministic with respect to external behavior: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in
\code{(call~f1()~||~call~f2())},
the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-inspired synchronous languages enforces an arbitrary 
execution order for statements in multiple lines of execution to achieve 
intra-reaction determinism
(%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}%                  % 2010
).
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

\begin{figure}[b]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are safe}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are unsafe}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}
